---
title: STAT 457 Homework 02
author: Martha Eichlersmith
date: 2019-10-07
output:
  pdf_document:
    fig_caption: yes
header-includes:
  - \usepackage{color}
  - \usepackage{mathtools}
---
```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
library(ggplot2)
library(readr)
library(gridExtra)
library(grid)
library(png)
library(downloader)
library(grDevices)
library(latex2exp)
library(knitr)
library(leaps)
library(directlabels)
library(diffusr)
library(MASS)
library(invgamma)
library(condMVNorm)
```

```{r, echo=FALSE}
#FUNCTION FOR PROBLEMS 0 & 1
#Sinlge Random Walk - Discrete (output is each step)
Single_Disc_Walk_2d <-function(n_steps) 
{ rw <- matrix(0, ncol = 2, nrow = n_steps) 
  indx <- cbind(seq(n_steps), sample(c(1, 2), n_steps, TRUE))  
  rw[indx] <- sample(c(-1, 1), n_steps, TRUE) 
  rw[1,1] <- 0 
  rw[1,2] <- 0
  rw[,1] <- cumsum(rw[, 1]) # cumsum the columns 
  rw[,2] <- cumsum(rw[, 2]) # cumsum the columns 
  return(as.data.frame(rw)) } # return values of each step in random walk 
#Multiple Discrete Random Walks, output is coordinates to end points 
Disc_Walk_2d <- function(n_steps, n_walks){
  coord <- matrix(0, ncol=2, nrow=n_walks) 
for (i in 1:n_walks){
 rw <- matrix(0, ncol = 2, nrow = n_steps) 
  indx <- cbind(seq(n_steps), sample(c(1, 2), n_steps, TRUE))  
  rw[indx] <- sample(c(-1, 1), n_steps, TRUE) 
  rw[1,1] <- 0 
  rw[1,2] <- 0
  rw[,1] <- cumsum(rw[, 1]) # cumsum the columns 
  rw[,2] <- cumsum(rw[, 2]) # cumsum the columns 
  coord[i,1] <- rw[n_steps,1]
  coord[i,2] <- rw[n_steps,2] }
  return(as.data.frame(coord)) }  # return end coord for n_walks
#Multiple Continuous  Random Walks, output is coordinates to end points 
Cont_Walk_2d <- function(n_steps, n_walks){
jump <- 1
x<- array(0, dim=n_walks)
y<- array(0, dim=n_walks)  
for(i in 1:n.walks){
  theta <- runif(n_steps, 0, 2*pi)
  dx <- jump*cos(theta)
  dy <- jump*sin(theta)
  x[i] <- sum(dx)
  y[i] <- sum(dy) }
coord <- data.frame(x,y)
return(coord) }
#calculate the percetage of end points in each quadrant, output is vector of percentages
Quad_pct <- function(coord){
Q1.points <- subset(coord, coord[1] >= 0 & coord[2] >=0)
Q2.points <- subset(coord, coord[1] <  0 & coord[2] > 0)
Q3.points <- subset(coord, coord[1] <= 0 & coord[2] <=0)
Q4.points <- subset(coord, coord[1] >  0 & coord[2] < 0)
Q1.pct <- nrow(Q1.points) / n.walks
Q2.pct <- nrow(Q2.points) / n.walks
Q3.pct <- nrow(Q3.points) / n.walks
Q4.pct <- nrow(Q4.points) / n.walks
prop<-c(Q1.pct, Q2.pct, Q3.pct, Q4.pct)
return(prop)}
```

## Problem 0.  
```{r, fig.width=10, fig.height=6, echo=FALSE}
set.seed(010000) #Homework 01 | Problem 00 | Part 00
walk <- Single_Disc_Walk_2d(50)
ggplot(aes(x=walk[,1], y=walk[,2]),data=walk) + ggtitle("Graph a random walk")+
  geom_path(color="blue", size=1) +
  geom_hline(yintercept=0, linetype="dashed") + 
  geom_vline(xintercept=0, linetype="dashed") + 
  geom_point(x=0, y=0, color="darkgreen", size=5) +
  geom_point(x=walk[nrow(walk),1], y=walk[nrow(walk),2], color="red", size=5) +
  theme(axis.title.y=element_blank(), axis.title.x=element_blank()    )
```

## Problem 1a.  
Simulate 10,000 random walks of length 50 in $\mathbb{R}^2$ starting at (0, 0). Compute the proportion of walks which end up in each of the four quadrants.  What values would you expect for these proportions? Do the observed proportion vary significantly from the expected values?  
```{r, echo=FALSE}
set.seed(010101) #Homework 01 | Problem 01 | Part 01
n.steps <- 50
n.walks <- 10000
coord <- Cont_Walk_2d(n.steps, n.walks)
quad.pct <- Quad_pct(coord)
quad.pct
chisq.test(quad.pct)
```
You would expect each quadrant to have approximately 25% of end points for the random walks.  The observed proportions do not vary significantly from the expected value.  
  
\newpage  
## Problem 1b.  
Repeat 1a using walks of length 500  
```{r, echo=FALSE}
set.seed(010102) #Homework 01 | Problem 01 | Part 02
n.steps <- 500
n.walks <- 10000
coord <- Cont_Walk_2d(n.steps, n.walks)
quad.pct <- Quad_pct(coord)
quad.pct
chisq.test(quad.pct)
```
Similar to 1a, the percetnages are very close to 0.25.  

## Problem 1c.  
Repeat 1a and 1b using random walks on a lattice in $\mathbb{R}^2$.  
```{r, echo=FALSE}
set.seed(010103) #Homework 01 | Problem 01 | Part 03.
n.steps <- 50
n.walks <- 10000
coord <- Disc_Walk_2d(n.steps, n.walks)
quad.pct <- Quad_pct(coord)
quad.pct
chisq.test(quad.pct)

set.seed(010104) #Homework 01 | Problem 01 | Part 04 .
n.steps <- 500
n.walks <- 10000
coord <- Disc_Walk_2d(n.steps, n.walks)
quad.pct <- Quad_pct(coord)
quad.pct
chisq.test(quad.pct)
```
Both of these situations have percentages that are close to 0.25.  It is interesting to point out that the $\chi^2$ values are larger for the lattice random walks rather than the continous random walks.  

\newpage  
## Problem 2.  
From each of the following distributions: (i) draw 10,000 deviates; (ii) compute the sample average and sample SD and compare these numbers to the true values; (iii) draw a histogram of the deviates along with the plot of the true density.  
Binomial(14, .90); Beta(.5, .5); Gamma(12, 2); Inverse.Gamma(12, 2); $\chi^2$ on 2 df; $\chi^{-2}$ on 2 df.  

### Problem 2a: Binomial(14, .9) 
```{r, fig.width=12, fig.height=6, message=FALSE, echo=FALSE}
set.seed(010201) #Homework 01 | Problem 02 | Part 01
r <- 5 #rounding decimals 
binom.fun <- function(d, n, p){
dist <- paste("Binomal - ", d, "devaites")  
X <- rbinom(d, n, p) 
X.df <- data.frame(X=X)
Name <- c("Mean", "SD")
sample <- c(round(mean(X), r), round(sd(X), r))
true <- c(round(n*p,r), round(sqrt(n*p*(1-p)),r))
compare <- data.frame("Name"=Name, "Sample"=sample, "True"=true)
plot <- ggplot(X.df, aes(X)) + geom_histogram(aes(y=..density..)) + 
  ggtitle(dist)+
  xlab("") + ylim(0, 1.5)+
  annotation_custom(tableGrob(compare), xmin=7, xmax=10, ymin=1) +
  stat_function(
  fun = dnorm, #normal approx to binom
  args = list(mean= n*p, sd= sqrt(n*p*(1-p))),
  lwd = 1,
  linetype="dashed",
  col = 'blue'
  ) 
return(plot)}

binom1 <- binom.fun(10000, 14, .9)
binom2 <- binom.fun(100000, 14, .9)
binom3 <- binom.fun(1000000, 14, .9)
grid.arrange(binom1, binom2, binom3, nrow = 1)
```

### Problem 2b: Beta(.5, .5)  
```{r, fig.width=12, fig.height=6, message=FALSE, warnings=FALSE, error=FALSE, echo=FALSE}
set.seed(010202) #Homework 01 | Problem 02 | Part 02
r <- 5 #rounding decimals 

beta.fun <- function(d, a, b){
dist <- paste("Beta - ", d, "devaites")  
X <- rbeta(d, a, b) 
X.df <- data.frame(X=X)
true <- c(round(a / (a + b), r), round(sqrt(a*b / ((a+b)^2 * (a + b +1))), r))
Name <- c("Mean", "SD")
sample <- c(round(mean(X), r), round(sd(X), r))
compare <- data.frame("Name"=Name, "Sample"=sample, "True"=true)
plot <- ggplot(X.df, aes(X)) + geom_histogram(aes(y=..density..)) + 
  ggtitle(dist)+
  xlab("") + 
  coord_cartesian(ylim=c(0, 2.5))+
  annotation_custom(tableGrob(compare), ymin=1) +
  stat_function(
  fun = dbeta, 
  args = list(shape1=a, shape2=0.5),
  lwd = 1,
  linetype="dashed",
  col = 'blue'
  ) 
return(plot)}

beta1 <- beta.fun(10000, .5, .5) 
beta2 <- beta.fun(100000, .5, .5)
beta3 <- beta.fun(1000000, .5, .5)
grid.arrange(beta1, beta2, beta3, nrow = 1)
```

### Problem 2c: Gamma(12, 2)  
```{r, fig.width=12, fig.height=6, message=FALSE, echo=FALSE}
set.seed(010203) #Homework 01 | Problem 02 | Part 03
r <- 5 #rounding decimals 

gamma.fun <- function(d, a, b){
dist <- paste("Gamma - ", d, " devaites") 
X <- rgamma(d, a, b) 
X.df <- data.frame(X=X)
true <- c(round(a/b, r), round(sqrt(a/(b^2)), r))
Name <- c("Mean", "SD")
sample <- c(round(mean(X), r), round(sd(X), r))
compare <- data.frame("Name"=Name, "Sample"=sample, "True"=true)
plot <- ggplot(X.df, aes(X)) + geom_histogram(aes(y=..density..)) + 
  ggtitle(dist)+
  xlab("") +
  annotation_custom(tableGrob(compare), ymin=.1, xmin=5) +
  stat_function(
  fun = dgamma,
  args = list(shape=a, rate=b, log=FALSE),
  lwd = 1,
  linetype="dashed",
  col = 'blue'
  ) 
return(plot)}

gamma1 <- gamma.fun(10000, 12, 2)
gamma2 <- gamma.fun(100000, 12, 2)
gamma3 <- gamma.fun(1000000, 12, 2)
grid.arrange(gamma1, gamma2, gamma3, nrow = 1)
```

### Problem 2d: Inverse.Gamma(12, 2) 
```{r, fig.width=12, fig.height=6, message=FALSE, echo=FALSE}
set.seed(010204) #Homework 01 | Problem 02 | Part 04
r <- 5 #rounding decimals 

invgamma.fun <- function(d, a, b){
dist <- paste("Inverse Gamma - ", d, " devaites") 
X <- rinvgamma(d, a, b) 
X.df <- data.frame(X=X)
true <- c(round(b/(a -1), r), round(sqrt( b^2 / ((a - 1)^2 * (a -2))), r))
Name <- c("Mean", "SD")
sample <- c(round(mean(X), r), round(sd(X), r))
compare <- data.frame("Name"=Name, "Sample"=sample, "True"=true)
plot <- ggplot(X.df, aes(X)) + geom_histogram(aes(y=..density..)) + 
  ggtitle(dist)+
  xlab("") +
  #ylim(0, 1.5)+
  annotation_custom(tableGrob(compare), ymin=5, xmin=.2) +
  stat_function(
  fun = dinvgamma, 
  args = list(shape=a, rate=b, log=FALSE),
  lwd = 1,
  linetype="dashed",
  col = 'blue'
  ) 
return(plot)}

invgam1 <- invgamma.fun(10000, 12, 2)
invgam2 <- invgamma.fun(100000, 12, 2)
invgam3 <- invgamma.fun(100000, 12, 2)
grid.arrange(invgam1, invgam2, invgam3, nrow = 1)
```

 
### Problem 2e: $\chi^2$ on 2 df  
```{r, fig.width=12, fig.height=6, message=FALSE, echo=FALSE}
set.seed(010205) #Homework 01 | Problem 02 | Part 05
r <- 5 #rounding decimals 

chisq.fun <- function(d, df){
dist <- paste("Chi-Squared - ", d, " devaites")
X <- rchisq(d, df) 
X.df <- data.frame(X=X)
true <- c(round(df, r), round(sqrt(2*df), r))
Name <- c("Mean", "SD")
sample <- c(round(mean(X), r), round(sd(X), r))
compare <- data.frame("Name"=Name, "Sample"=sample, "True"=true)
plot <- ggplot(X.df, aes(X)) + geom_histogram(aes(y=..density..)) + 
  ggtitle(dist)+
  xlab("") +
  #ylim(0, 1.5)+
  annotation_custom(tableGrob(compare), ymin=.3, xmin=2) +
  stat_function(
  fun = dchisq, 
  args = list(df=df),
  lwd = 1,
  linetype="dashed",
  col = 'blue'
  ) 
return(plot)}

chisq1 <- chisq.fun(10000, 2)
chisq2 <- chisq.fun(100000, 2)
chisq3 <- chisq.fun(1000000, 2)
grid.arrange(chisq1, chisq2, chisq3, nrow = 1)
```

### Problem 2f: $\chi^{-2}$ on 2 df
```{r, fig.width=12, fig.height=6, message=FALSE, echo=FALSE}
set.seed(010206) #Homework 01 | Problem 02 | Part 06
r <- 5 #rounding decimals 

invchisq.fun <- function(d, df){
dist <- paste("Inverse Chi-Squared - ", d, " devaites")
X <- rinvchisq(d, df) 
X.df <- data.frame(X=X)
true <- c("DNE when df <=2", "DNE when df <=4")
Name <- c("Mean", "SD")
sample <- c(round(mean(X), r), round(sd(X), r))
compare <- data.frame("Name"=Name, "Sample"=sample, "True"=true)
plot <- ggplot(X.df, aes(X)) + geom_histogram(aes(y=..density..)) + 
  ggtitle(dist)+
  xlab("") +
  #ylim(0, 1.5)+
  annotation_custom(tableGrob(compare)) +
  stat_function(
  fun = dinvchisq,
  args = list(df=df),
  lwd = 1,
  linetype="dashed",
  col = 'blue'
  ) 
return(plot)}

invchi1 <- invchisq.fun(10000, 2)
invchi2 <- invchisq.fun(100000, 2)
invchi3 <- invchisq.fun(1000000, 2)
grid.arrange(invchi1, invchi2, invchi3, nrow = 1)
```

\newpage  
## Problem 3a.  
Draw a sample size of 10,000 from the trivariate normal distribution with mean (0, 0, 0) and variance-covariance matrix:  
$$
\begin{bmatrix} 
1 & 4.5 & 9 \\ 
4.5 & 25 & 49 \\
9 & 49 & 100 
\end{bmatrix}
$$

## Problem 3b.  
Draw a histogram of the $X_1$ deviates, along with the true marginal.  Compute the sample average and sample SD and compare these numbers to the true values.  
```{r, fig.width=12, fig.height=6, message=FALSE, echo=FALSE}
set.seed(010301)  #Homework 01 | Problem 03 | Part 01
r <- 5 #rounding decimals 
mu <- c(0,0,0)
sig <- matrix(c(1,4.5,9,4.5,25,49,9,49,100),nrow=3, ncol=3) 
mvn.fun <- function(d, mu, sig){
dist <- paste("Multivariate Normal - ", d, "devaites")  
MVN.data <-mvrnorm(d, mu, sig, tol = 1e-6, empirical = FALSE, EISPACK = FALSE)
MVN.data<-data.frame(MVN.data, nrow = d, ncol = 3)
colnames(MVN.data)<-c("X1","X2","X3")
X1 <- MVN.data$X1
true <- c(0, 1)
Name <- c("Mean", "SD")
sample <- c(round(mean(X1), r), round(sd(X1), r))
compare <- data.frame("Name"=Name, "Sample"=sample, "True"=true)
plot <- ggplot(MVN.data, aes(X1)) + geom_histogram(aes(y=..density..)) + 
  ggtitle(dist)+xlab("") +
  annotation_custom(tableGrob(compare),xmax=0,ymin =0.05) +
  stat_function(fun = dnorm,args = list(mean=0, sd=1),lwd = 1,
                linetype="dashed", col = 'blue')  
return(plot)}

mvn1 <- mvn.fun(10000, mu, sig)
mvn2 <- mvn.fun(100000, mu, sig)
mvn3 <- mvn.fun(1000000, mu, sig)
grid.arrange(mvn1, mvn2, mvn3, nrow = 1)
```


\newpage  
## Problem 3c.  
Draw a sample size of 10,000 from the conditional distribution $p(X_1 \mid X_2, X_3)$, take $X_2 = X_3 = 1$.  Compute the sample mean and sample SD and compare these numbers to the true values.  Draw a histogram of the simulate values, along with the true conditional distribution.  
```{r, fig.width=12, fig.height=6, message=FALSE, echo=FALSE}
set.seed(010303)  #Homework 01 | Problem 03 | Part 03
mu <- c(0, 0, 0)
sig <- matrix(c(1,4.5,9,4.5,25,49,9,49,100),nrow=3, ncol=3) 
cmvn.fun <- function(d, mu, sig){
X<-rcmvnorm(d, mu, sig, dependent.ind = c(1), given.ind = c(2,3), X.given = c(1,1) )
X<-matrix(X, nrow = d, ncol = 1)
X.df <- data.frame(X=X)
x.g<- matrix(c(1,1), ncol=1)
sig.11 <-sig[1,1] 
sig.1g <- t(as.matrix(sig[1, 2:3]))
sig.g1 <- t(sig.1g)
sig.gg<- sig[2:3, 2:3]
sig.gg_inverse<-solve(sig.gg)
cond.mean<- sig.1g %*% sig.gg_inverse %*% x.g
cond.sd <- sqrt(sig.11 - sig.1g %*% sig.gg_inverse %*% sig.g1)
dist <- paste("Conditional Multivariate Normal - ", d, " deviates") 
r <- 5 #decimals to round to 
true <- c(round(cond.mean, r), round(cond.sd, r))
Name <- c("Mean", "SD")
sample <- c(round(mean(X), r), round(sd(X), r))
compare <- data.frame("Name"=Name, "Sample"=sample, "True"=true)
plot <- ggplot(X.df, aes(X)) + geom_histogram(aes(y=..density..)) + 
  ggtitle(dist)+  xlab("") +
  annotation_custom(tableGrob(compare),ymin =0.5) +
  stat_function(fun = dnorm, args = list(mean=cond.mean, sd=cond.sd), lwd = 1,
                linetype="dashed", col = 'blue'  ) 
return(plot)}

cmvn1 <- cmvn.fun(10000, mu, sig)
cmvn2 <- cmvn.fun(100000, mu, sig)
cmvn3 <- cmvn.fun(1000000, mu, sig)
grid.arrange(cmvn1, cmvn2, cmvn3, nrow = 1)
```

We want to find  expected value and sd of $X_1 | X_2 = X_3 = 1$. Let $\pmb{X}_g = (X_2, X_3)$

 $$
 \begin{aligned} 
 X_1 |\pmb{X}_g &\sim N( \mu_1 + \Sigma_{1g}\Sigma_{gg}^{-1}(\pmb{x}_g -\pmb{\mu}_g), \Sigma_{11} - \Sigma_{22}^{-1} \Sigma_{12}^T)
 \\ \quad \\
\mathbb{E} \left[ X_1 \mid \pmb{X}_g \right] & = \mu_1 + \Sigma_{1g}\Sigma_{gg}^{-1}(\pmb{x}_g -\pmb{\mu}_g) 
\\
& = 0 + 
\begin{bmatrix} 4.5 & 9 \end{bmatrix} 
\cdot \begin{bmatrix} 25 & 49 \\ 49 & 100 \end{bmatrix}^{-1} 
\left( 
\begin{bmatrix} 1 \\ 1 \end{bmatrix} 
- 
\begin{bmatrix} 0 \\ 0 \end{bmatrix} 
\right)
\\
& = 0.1363636
\\
\quad \\
\text{s.d.} ( X_1 \mid \pmb{X}_g ) & = 0.4264014
 \end{aligned}
 $$



## Problem 3d.  
Use the sample size of 10,000 from the joint trivariate distribution to obtain he conditional mean and SD of $X_1$ given $X_2 = X_3 = 1$, i.e. look at the triples whose second and third components are within $\epsilon$ of 1.  How do these values compare to the true mean and SD for various values of $\epsilon$?  
```{r, echo=FALSE}
set.seed(010304)  #Homework 01 | Problem 03 | Part 04
d <-10000
mu <- c(0, 0, 0)
sig <- matrix(c(1,4.5,9,4.5,25,49,9,49,100),nrow=3, ncol=3) 
MVN.data <-mvrnorm(d, mu, sig, tol = 1e-6, empirical = FALSE, EISPACK = FALSE)
MVN.data<-data.frame(MVN.data, nrow = d, ncol = 3)
colnames(MVN.data)<-c("X1","X2","X3")
X1 <- MVN.data$X1
ep <- function(ep){
x2x3<-subset(MVN.data, abs(MVN.data$X2-1) < ep & abs(MVN.data$X3-1) < ep)
epx1<-x2x3[,1]
return(paste("Epsilon of ", ep, ": mean = ", mean(epx1), " and sd = ", sd(epx1)))
}
sapply(c(0.1, 0.5, 0.25), ep)
```

