---
#title: STAT 457 Delinquency Problem 
#author: Martha Eichlersmith
output:
  pdf_document:
    fig_caption: yes
header-includes:
  - \usepackage{color}
  - \usepackage{mathtools}
  - \usepackage{amsbsy} #bold in mathmode
  - \usepackage{nicefrac} # for nice fracs 
  - \usepackage{booktabs}
  - \usepackage{geometry}
  - \usepackage{caption} #to remove automatic table name and number - \captionsetup[table]{labelformat=empty}, put code under ---
  - \usepackage{lastpage} #for pageref*{LastPage} - to get total # of pages
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[L]{STAT 457 Fall 2019 \\ Delinquency Problem}
  - \fancyhead[R]{Martha Eichlersmith \\ Page \thepage\ of\ \pageref*{LastPage}}  
  - \setlength{\headheight}{22.5pt} #to remove \fancyhead error for head height
#left-even pages (LE) right-odd pages (RE)
geometry: "left=0.75in,right=0.75in,top=1.1in,bottom=1in" 

---

\captionsetup[table]{labelformat=empty}
```{r setup, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
library(ggplot2) #ggplot
library(readr) #import CSV
library(gridExtra) #organize plots
library(grid) #organize plots
library(latex2exp) #latex in ggplot titles 
library(gtable) #for tablegrob functions 
#library(kableExtra) #for kable functions
library(dplyr) #for piping 
library(MCMCpack) #for dirichelt
knitr::opts_chunk$set(fig.width = 10, fig.height = 4)
knitr::opts_chunk$set(echo=FALSE)
decimal <- function(x, k) trimws(format(round(x, k), nsmall=k))
dec <- 3
#knitr::opts_chunk$set(echo=FALSE) #using knitr for this option but don't have to load 
```

# Is my little brother a delinquent? 

```{r Counts}
alpha <-  c(127, 123, 93, 17) #most delinquent
gamma <-  c(345, 209, 158, 65)  #least delinquent

table <- rbind( alpha, gamma)
rownames(table) <- c("Most Delinquent", "Least Delinquent")
colnames(table) <- c("Oldest", "In-between", "Youngest", "Only Child")

knitr::kable(table, booktabs=TRUE, 'latex', digits=dec, caption="Raw Data") %>%
  kableExtra::kable_styling(latex_options="hold_position" ) 
```
Using the above data, we want to know if birth order affects delinquency in boys.  Looking at the raw data row proportions it appears that the oldest and only-child sons have a lower delinquency level, where as middle and youngest sons have a higher delinquency levels.   
```{r func_tableparam}
#FUNCTION: PUT PARAMETERS INTO TABLE
func_tableparam <- function(parameters, name){
diff <- parameters[1,]- parameters[2,]
table <- rbind( parameters, diff)
rownames(table) <- c("Most Delinquent", "Least Delinquent", "Difference (Most - Least)")
colnames(table) <- c("Oldest", "In-between", "Youngest", "Only Child")
knitr::kable(table, booktabs=TRUE, 'latex', digits=dec, caption=paste(name)) %>%
  kableExtra::kable_styling(latex_options="hold_position" ) 
}

func_tableparam(rbind(alpha/sum(alpha), gamma/sum(gamma)), "Row Proportions")
```


\newpage 
## Bayesian


```{r func_values}
#FUNCTION: RETURN IT, MEAN, CI, AND PVALUE 
func_values <- function(vec){
  it <- length(vec)
  CI.val <- decimal(quantile(vec, c(0.025, 0.975)), dec)
  CI.95 <- paste("(", paste(CI.val, collapse=", "), ")")
  mean <- decimal(mean(vec), dec)
  pval <- decimal(min(length(vec[vec > 0])/it, 1 - length(vec[vec > 0])/it), dec)
  vec <- c(it, mean, CI.95, pval)
  vec
}
```
```{r func_delinq}
#FUNCTION: SIMIULATE VALUES 
func_delinq <- function(it, alpha, gamma){
set.seed(050104)
prop.alpha <- rdirichlet(it, alpha)
prop.gamma <- rdirichlet(it, gamma)
diff <- cbind( #most - least
   prop.alpha[,1] - prop.gamma[,1] #diff for Oldest
  ,prop.alpha[,2] - prop.gamma[,2] #diff for in-between
  ,prop.alpha[,3] - prop.gamma[,3] #diff for youngest
  ,prop.alpha[,4] - prop.gamma[,4] #diff for only-child
   )
apply(diff, 2, func_values)
}
```
```{r func_table}
#FUNCTION: put into a table 
func_table <- function(i, name, it.1, it.2, it.3){
table <- cbind(it.1[,i], it.2[,i] , it.3[,i])
rownames(table) <- c("Iterations", "Mean", "95% CI", "pval")
title <- paste("Difference in Proportion for", name, ": Most Delinquent - Least Delinquent")

knitr::kable(table, booktabs=TRUE, 'latex', digits=dec, caption=paste(title)) %>%
  kableExtra::kable_styling(latex_options="hold_position" ) 
}
```
```{r func_chisq}
#FUNCTION: CHI SQUARE - COMBINING P-VALUES 
func_pval.chisq <- function(pvals){
  x <- -2*sum(log(as.numeric(pvals)))
  combined.pval <-  1- pchisq(x, 2*length(pvals))
}
```
```{r func_combinedpval}
#FUNCTION: COMBINE THE P-VALUES INTO TABLE 
func_combinedpval <- function(it.1, it.2, it.3){
pvals.matrix <- cbind(  as.numeric(it.1[4,])
                      , as.numeric(it.2[4,])
                      , as.numeric(it.3[4,])
)

combined <- apply(pvals.matrix, 2, func_pval.chisq)
combined.pvals <- rbind(it.vec, decimal(combined, 10))
rownames(combined.pvals) <- c("Iterations", "Combined p-value")

title <- paste("Combined p-values:","$X = -2\\sum_{i=1}^4(p_i) \\sim \\chi^2_{df=8}$")
knitr::kable(combined.pvals, booktabs=TRUE, 'latex', caption=paste(title)) %>%
  kableExtra::kable_styling(latex_options="hold_position" ) 
}
```
```{r bayesRESULTS}

it.vec <- c(1e04, 1e05, 1e06)
it1 <- func_delinq(it.vec[1], alpha, gamma)
it2 <- func_delinq(it.vec[2], alpha, gamma)
it3 <- func_delinq(it.vec[3], alpha, gamma)

func_table(1, "Oldest", it1, it2, it3)
func_table(2, "In-Between", it1, it2, it3)
func_table(3, "Youngest", it1, it2, it3)
func_table(4, "Only", it1, it2, it3)
func_combinedpval(it1, it2, it3)
```
Using simulations, the results support the hypothesis that delinquency levels are higher in middle and younger sons but lower in oldest and only sons.  

\newpage 
## Frequentest Tests
### Proportion Test for Specific Birth Order 

```{r func_proptest}
func_proptest <- function(c, alt, name){
table2x2 <- matrix(c(alpha[c], sum(alpha), gamma[c], sum(gamma)), nrow=2, byrow=TRUE)
test <- prop.test(table2x2, alternative=alt)
X.squared <- decimal(test$statistic, dec)
p.value <- decimal(test$p.value, dec)
alternative <- test$alternative
results <- cbind(X.squared, p.value, alternative)
rownames(results) <-c("")
title <- paste("Difference in Proportion for", name, ": Most Delinquent - Least Delinquent")
knitr::kable(results, booktabs=TRUE, 'latex', caption=paste(title)) %>%
  kableExtra::kable_styling(latex_options="hold_position" ) 
}
```

```{r freqRESULTS1}
func_proptest(1, 'less',"Only")
func_proptest(2, 'greater',"Only")
func_proptest(3, 'greater',"Only")
func_proptest(4, 'less',"Only")
```



### Chi-Sq test for Independence 
```{r freqRESULTS2}
parameters <- rbind(alpha, gamma)
chisq.test(parameters)
```

Using traditional frequentest test, the results are the same: delinquency levels are higher in middle and younger sons but lower in oldest and only sons.  

## Conclusion
Using the data provided, there is strong evidence that younger sons are more likely to be a delinquent.  So even though my brother is not is a delinquent, there is a higher probably that he will become one.  


```{r}
#PRINTING THE CODE
#knitr::stitch("HW06.Rmd") to go to latex
knitr::stitch(   script="STAT457-DelinquencyProblem.Rmd"  , system.file("misc", "knitr-template.Rhtml", package="knitr")) #code to HTML
```